### 2020-09-25 submission/manuscript.tex
Add edits from Pat

Differences in CDI mouse model studies have been attributed to
intestinal microbiota variation across sources. For example, researchers
using the same clindamycin treatment and C57BL/6 mice had different
\emph{C. difficile} outcomes, one having sustained colonization (32),
while the other had transient colonization (18), despite both using
\emph{C. difficile} VPI 10643. Baseline differences in the microbiota
composition have been hypothesized to partially explain the differences
in colonization outcomes and overall susceptibility to \emph{C.
difficile} after treatment with the same antibiotic (13, 31). When we
treated mice from 6 different sources with clindamycin and challenged
them with \emph{C. difficile} 630, we found microbiota variation across
sources impacted colonization outcomes, but not susceptibility. A
previous study with \emph{C. difficile} identified an endogenous
protective \emph{C. difficile} strain LEM1 that bloomed after antibiotic
treatment in mice from Jackson or Charles River Laboratories, but not
Taconic that protected mice against the more toxigenic \emph{C.
difficile} VPI10463 (26). Given that we obtained mice from the same
vendors, we checked all mice for endogenous \emph{C. difficile} by
plating stool samples that were collected after clindamycin treatment.
However, we did not identify any endogenous \emph{C. difficile} strains
prior to challenge, suggesting there were no endogenous protective
strains in the mice we received and other bacteria mediated the
variation in \emph{C. difficile} colonization across sources. The
\emph{C. difficile} strain used could also be contributing to the
variation in \emph{C. difficile} outcomes seen across different research
groups. For example, a group found differential colonization outcomes
after clindamycin treatment, with \emph{C. difficile} 630 and M68
infections eventually becoming undetectable while strain BI-7 remained
detectable up to 70 days post-treatment (44). One study limitation is
that we only used female mice. Sex has been shown to influence
microbiota variation in mice (45), so we used female mice to reduce this
confounding variable and also match the sex used in previous CDI studies
that administered clindamycin to mice (32, 32, 44, 46). The bacterial
perturbations induced by clindamycin treatment have been well
characterized and our findings agree with previous CDI mouse model work
demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae} were
associated with \emph{C. difficile} susceptibility and
\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
resistance (19, 21, 32, 33, 43, 44, 46, 47). While we have demonstrated
that susceptibility is uniform across sources of mice after clindamycin
treatment, there could be different outcomes for either susceptibility
or clearance in the case of other antibiotic treatments.

@@

\textbf{(v) 16S rRNA gene sequence analysis.} mothur (v. 1.43) was used
to process all sequences (71) with a previously published protocol (69).
Reads were combined and aligned with the SILVA reference database (72).
Chimeras were removed with the VSEARCH algorithm and taxonomic
assignment was completed with a modified version (v16) of the Ribosomal
Database Project reference database (v11.5) (73) with an 80\% confidence
cutoff. Operational taxonomic units (OTUs) were assigned with a 97\%
similarity threshold using the opticlust algorithm (74). Based on the
mock communities, our overall sequencing error rate was 0.0112\% and all
water controls had less than 1000 sequences (range: 18-875). To account
for uneven sequencing across samples, samples were rarefied to 5,437
sequences 1,000 times for alpha and beta diversity analyses, and a
single time to generate relative abundances for model training. PCoAs
were generated based on the Yue and Clayton measure of dissimilarity
(\(\theta_{YC}\)) distances. Permutational multivariate analysis of
variance (PERMANOVA) was performed on mothur-generated \(\theta_{YC}\)
distance matrices with the adonis function in the vegan package (75) in
R (76).

\textbf{(vi) Classification model training and evaluation.} Models were
generated based on mice that were categorized as either cleared or
colonized 7 days post-infection and had sequencing data from the
baseline (day -1), post-clindamycin (day 0), and post-infection (day 1)
timepoints of the experiment. Input bacterial community relative
abundance data at the OTU level from the baseline, post-clindamycin, and
1-day post-infection timepoints was used to generate 3 classification
models that predicted \emph{C. difficile} colonization status 7 days
post-infection. The L2-regularized logistic regression models were
trained and tested using the caret package (77) in R as previously
described (78) with the exception that we used 60\% training and 40\%
testing data splits for testing of the held out test data to measure
model performance and repeated k-fold cross-validation of the training
data to select the best cost hyperparameter. The modified training to
testing ratio was selected to accommodate the small number of samples in
the dataset. Code was modified from
\url{https://github.com/SchlossLab/ML_pipeline_microbiome} to update the
classification outcomes and change the data split ratios. The modified
repository to regenerate our modeling analysis is available at
\url{https://github.com/tomkoset/ML_pipeline_microbiome}.

\textbf{(vii) Statistical analysis.} All statistical tests were
performed in R (v 4.0.2) (76). The Kruskal-Wallis test was used to
analyze differences in \emph{C. difficile} CFU, mouse weight change, and
alpha diversity across sources with a Benjamini-Hochberg correction for
testing multiple timepoints, followed by pairwise Wilcoxon comparisons
with Benjamini-Hochberg correction. For taxonomic analysis and
generation of logistic regression model input data, \emph{C. difficile}
(OTU 20) was removed. Bacterial relative abundances that varied across
sources at the OTU level were identified with the Kruskal-Wallis test
with Benjamini-Hochberg correction for testing all identified OTUs,
followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg
correction. The Wilcoxon rank sum test was used to test for OTUs that
differed between experiments within the Schloss, Young, and Envigo
sources with Benjamini-Hochberg correction for testing all identified
OTUs. OTUs impacted by clindamycin treatment were identified using the
paired Wilcoxon signed rank test with matched pairs of mice samples from
day -1 and day 0. To determine whether classification models had better
performance (test AUROCs) than random chance (0.5), we used the
one-sample Wilcoxon signed rank test. To examine whether there was an
overall difference in predictive performance across the 3 classification
models we used the Kruskal-Wallis test followed by pairwise Wilcoxan
comparisons with Benjamini-Hochberg correction for multiple hypothesis
testing. The tidyverse package (v 1.3.0) was used to wrangle and graph
data (79).

@@

\leavevmode\hypertarget{ref-Vegan2018}{}%
75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
Wagner H. 2018. Vegan: Community ecology package.

\leavevmode\hypertarget{ref-r_citation_2018}{}%
76. R Core Team. 2018. R: A language and environment for statistical
computing. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\hypertarget{ref-Kuhn2008}{}%
77. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
Journal of Statistical Software 28.

\leavevmode\hypertarget{ref-Topcuoglu2020}{}%
78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
framework for effective application of machine learning to
microbiome-based classification problems. mBio 11.

\leavevmode\hypertarget{ref-Tidyverse2019}{}%
79. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K,
Vaughan D, Wilke C, Woo K, Yutani H. 2019. Welcome to the tidyverse.
Journal of Open Source Software 4:1686.

@@

\includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that
influenced whether mice cleared \emph{C. difficile} by day 7.} A.
Post-clindamycin (day 0) relative abundance data for the 10 OTUs with
the highest rankings based on feature weights in the post-clindamycin
(day 0) classification model. Red font represents OTUs that correlated
with \emph{C. difficile} colonization and blue font represents OTUs that
correlated with clearance. Symbols represent the relative abundance data
for an individual mouse. Gray bars indicate the median relative
abundances for each source. The gray horizontal lines indicates the
limit of detection. B. Venn diagram that combines OTUs that were
important to the day -1, 0, and 1 classification models ( Fig. S4, Table
S14) and either overlapped with taxa that varied across sources at the
same timepoint, were impacted by clindamycin treatment, or both. Bold
OTUs were important to more than 1 classification model.

>>>>

Differences in CDI mouse model studies have been attributed to
intestinal microbiota variation across sources. For example, researchers
using the same clindamycin treatment and C57BL/6 mice had different
\emph{C. difficile} outcomes, one having sustained colonization (32),
while the other had transient colonization (18), despite both using
\emph{C. difficile} VPI 10643. Baseline differences in the microbiota
composition have been hypothesized to partially explain the differences
in colonization outcomes and overall susceptibility to \emph{C.
difficile} after treatment with the same antibiotic (13, 31). When we
treated mice from 6 different sources with clindamycin and challenged
them with \emph{C. difficile} 630, we found microbiota variation across
sources impacted colonization outcomes, but not susceptibility. A
previous study with \emph{C. difficile} identified an endogenous
protective \emph{C. difficile} strain LEM1 that bloomed after antibiotic
treatment in mice from Jackson or Charles River Laboratories, but not
Taconic that protected mice against the more toxigenic \emph{C.
difficile} VPI10463 (26). Given that we obtained mice from the same
vendors, we checked all mice for endogenous \emph{C. difficile} by
plating stool samples that were collected after clindamycin treatment.
However, we did not identify any endogenous \emph{C. difficile} strains
prior to challenge, suggesting there were no endogenous protective
strains in the mice we received and other bacteria mediated the
variation in \emph{C. difficile} colonization across sources. The
\emph{C. difficile} strain used could also be contributing to the
variation in \emph{C. difficile} outcomes seen across different research
groups. For example, a group found differential colonization outcomes
after clindamycin treatment, with \emph{C. difficile} 630 and M68
infections eventually becoming undetectable while strain BI-7 remained
detectable up to 70 days post-treatment (44). One study limitation is
that we only used female mice. Sex has been shown to influence
microbiota variation in mice (45), so we used female mice to reduce this
confounding variable and also match the sex used in previous CDI studies
that administered clindamycin to mice (32, 33, 44, 46). The bacterial
perturbations induced by clindamycin treatment have been well
characterized and our findings agree with previous CDI mouse model work
demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae} were
associated with \emph{C. difficile} susceptibility and
\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
resistance (19, 21, 32, 33, 43, 44, 46, 47). While we have demonstrated
that susceptibility is uniform across sources of mice after clindamycin
treatment, there could be different outcomes for either susceptibility
or clearance in the case of other antibiotic treatments.

@@

\textbf{(v) 16S rRNA gene sequence analysis.} mothur (v. 1.43) was used
to process all sequences (71) with a previously published protocol (69).
Reads were combined and aligned with the SILVA reference database (72).
Chimeras were removed with the VSEARCH algorithm and taxonomic
assignment was completed with a modified version (v16) of the Ribosomal
Database Project reference database (v11.5) (73) with an 80\% confidence
cutoff. Operational taxonomic units (OTUs) were assigned with a 97\%
similarity threshold using the opticlust algorithm (74). Based on the
mock communities, our overall sequencing error rate was 0.0112\% and all
water controls had less than 1000 sequences (range: 18-875). To account
for uneven sequencing across samples, samples were rarefied to 5,437
sequences 1,000 times for alpha and beta diversity analyses, and a
single time to generate relative abundances for model training. PCoAs
were generated based on the Yue and Clayton measure of dissimilarity
(\(\theta_{YC}\)) distances (75). Permutational multivariate analysis of
variance (PERMANOVA) was performed on mothur-generated \(\theta_{YC}\)
distance matrices with the adonis function in the vegan package (76) in
R (77).

\textbf{(vi) Classification model training and evaluation.} Models were
generated based on mice that were categorized as either cleared or
colonized 7 days post-infection and had sequencing data from the
baseline (day -1), post-clindamycin (day 0), and post-infection (day 1)
timepoints of the experiment. Input bacterial community relative
abundance data at the OTU level from the baseline, post-clindamycin, and
1-day post-infection timepoints was used to generate 3 classification
models that predicted \emph{C. difficile} colonization status 7 days
post-infection. The L2-regularized logistic regression models were
trained and tested using the caret package (78) in R as previously
described (79) with the exception that we used 60\% training and 40\%
testing data splits for testing of the held out test data to measure
model performance and repeated 2.5-fold cross-validation of the training
data to select the best cost hyperparameter. The modified training to
testing ratio was selected to accommodate the small number of samples in
the dataset. Code was modified from
\url{https://github.com/SchlossLab/ML_pipeline_microbiome} to update the
classification outcomes and change the data split ratios. The modified
repository to regenerate our modeling analysis is available at
\url{https://github.com/tomkoset/ML_pipeline_microbiome}.

\textbf{(vii) Statistical analysis.} All statistical tests were
performed in R (v 4.0.2) (77). The Kruskal-Wallis test was used to
analyze differences in \emph{C. difficile} CFU, mouse weight change, and
alpha diversity across sources with a Benjamini-Hochberg correction for
testing multiple timepoints, followed by pairwise Wilcoxon comparisons
with Benjamini-Hochberg correction. For taxonomic analysis and
generation of logistic regression model input data, \emph{C. difficile}
(OTU 20) was removed. Bacterial relative abundances that varied across
sources at the OTU level were identified with the Kruskal-Wallis test
with Benjamini-Hochberg correction for testing all identified OTUs,
followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg
correction. The Wilcoxon rank sum test was used to test for OTUs that
differed between experiments within the Schloss, Young, and Envigo
sources with Benjamini-Hochberg correction for testing all identified
OTUs. OTUs impacted by clindamycin treatment were identified using the
paired Wilcoxon signed rank test with matched pairs of mice samples from
day -1 and day 0. To determine whether classification models had better
performance (test AUROCs) than random chance (0.5), we used the
one-sample Wilcoxon signed rank test. To examine whether there was an
overall difference in predictive performance across the 3 classification
models we used the Kruskal-Wallis test followed by pairwise Wilcoxan
comparisons with Benjamini-Hochberg correction for multiple hypothesis
testing. The tidyverse package (v 1.3.0) was used to wrangle and graph
data (80).

@@

\leavevmode\hypertarget{ref-Yue2005}{}%
75. Yue JC, Clayton MK. 2005. A similarity measure based on species
proportions. Communications in Statistics - Theory and Methods
34:2123--2131.

\leavevmode\hypertarget{ref-Vegan2018}{}%
76. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
Wagner H. 2018. Vegan: Community ecology package.

\leavevmode\hypertarget{ref-r_citation_2018}{}%
77. R Core Team. 2018. R: A language and environment for statistical
computing. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\hypertarget{ref-Kuhn2008}{}%
78. Kuhn M. 2008. Building predictive models inRUsing thecaretPackage.
Journal of Statistical Software 28.

\leavevmode\hypertarget{ref-Topcuoglu2020}{}%
79. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
framework for effective application of machine learning to
microbiome-based classification problems. mBio 11.

\leavevmode\hypertarget{ref-Tidyverse2019}{}%
80. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K,
Vaughan D, Wilke C, Woo K, Yutani H. 2019. Welcome to the tidyverse.
Journal of Open Source Software 4:1686.

@@

\includegraphics{figure_5.pdf} \textbf{Figure 5. Bacteria that
influenced whether mice cleared \emph{C. difficile} by day 7.} A.
Post-clindamycin (day 0) relative abundance data for the 10 OTUs with
the highest rankings based on feature weights in the post-clindamycin
(day 0) classification model. Red font represents OTUs that correlated
with \emph{C. difficile} colonization and blue font represents OTUs that
correlated with clearance. Symbols represent the relative abundance data
for an individual mouse. Gray bars indicate the median relative
abundances for each source. The gray horizontal lines indicate the limit
of detection. B. Venn diagram that combines OTUs that were important to
the day -1, 0, and 1 classification models ( Fig. S4, Table S14) and
either overlapped with taxa that varied across sources at the same
timepoint, were impacted by clindamycin treatment, or both. Bold OTUs
were important to more than 1 classification model.
